---
title: "data.table and tidyverse"
author: "Jeremy Allen"
date: "`r format(Sys.time(), tz='America/New_York', usetz = TRUE)`"
output: html_document
editor_options: 
  chunk_output_type: inline
---

----

<br>

# Why Use data.table?

Nice feature list here: [data.table](https://rdatatable.gitlab.io/data.table/)  

<br>
<br>

```{r echo=TRUE, include=TRUE}

library(data.table)
library(tictoc)
```

Let's use fread() to quickly grab some online COVID data

```{r echo=TRUE, include=TRUE}

# get state data from The New York Times github
states <- fread(
  input = "https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv",
  key = c('state', 'date')
)

# get US national data from The New York Times github
us <- fread(
  input = "https://raw.githubusercontent.com/nytimes/covid-19-data/master/us.csv",
  key = 'date'
)

# get table of initial closure dates
#stay_at_home_table <- fread("https://raw.githubusercontent.com/jeremy-allen/covid-deaths/main/home_dates.csv")

```

## Set up a tibble and data.table: df and dt

```{r echo=TRUE, include=TRUE}

# we haven't loaded tidyverse yet, so let's do now
library(tidyverse, quietly = TRUE)

# reminder, R has a built-in dataset
states

# for convenience let's have a dt and df for comparing
dt <- states
df <- as_tibble(states)


```

## First look at data.table syntax

**dt[i,j,by]**  

What is dt[i,j,by]? Interpret this as: take dt, subset/reorder rows using i, then calculate j, grouped by by  

## I

```{r echo=TRUE, include=TRUE}

# i (think filtering by rows)
dt[state == "Georgia",]

```

#### Your turn! What happens here?

```{r echo=TRUE, include=TRUE}

dt[1:3]

```

#### Compared to here?

```{r echo=TRUE, include=TRUE}

df[1:3]

```

In data.table `dt[1:3]` filters for the first three rows. In base R and tidyverse, `df[1:3]` selects the first three columns.  

## J

```{r echo=TRUE, include=TRUE}

# j (think doing work on columns)
dt[, max(deaths)]

```

## By

```{r echo=TRUE, include=TRUE}

# by (think grouping your work by categories of a variable)
dt[, .(max_deaths = max(deaths)), by = state]

```

## tidyverse equivalents of I, J, and By

### I = filter()

```{r echo=TRUE, include=TRUE}

# i = filter
df %>% 
  filter(state == "Georgia")

```

### J = doing any work on columns

```{r echo=TRUE, include=TRUE}
# j = 
df %>% 
  pull(deaths) %>% 
  max()

```

### By = group_by() with summarise()

```{r echo=TRUE, include=TRUE}

# group_by() alone will error

# df %>% 
#   group_by(state) %>% 
#   max(deaths) # will error

# group_by() with summarise() is the tidyverse way

df %>% 
  group_by(state) %>% 
  summarise(deaths = max(deaths))

```

## Compare data.table and tidyverse methods

Using real-world data  

### tidyverse way  

```{r echo=TRUE, include=TRUE}

#---- TOTAL DEATHS BY STATE ----

# tidyverse way

df %>% 
  group_by(state) %>% 
  arrange(state, date) %>% 
  summarise(deaths_total = last(deaths)) %>% 
  ungroup() %>% 
  arrange(desc(deaths_total))

```

### data.table way  

```{r echo=TRUE, include=TRUE}

# data.table way

setkey(dt, state, date) # makes an ordered index of values

dt[, .(deaths_total = data.table::last(deaths)), by = state][order(-deaths_total)]

```

### Let's time them  

**Timed: the tidyverse way** 

```{r echo=TRUE, include=TRUE}

tictoc::tic("tidyverse way")
df %>% 
  group_by(state) %>% 
  arrange(state, date) %>% 
  summarise(deaths_total = last(deaths)) %>% 
  ungroup() %>% 
  arrange(desc(deaths_total))
tictoc::toc()

```

**Timed: the data.table way** 

```{r echo=TRUE, include=TRUE}

tictoc::tic("data.table way")
setkey(dt, state, date)
dt[, .(deaths_total = data.table::last(deaths)), by = state][order(-deaths_total)]
tictoc::toc()

```

### If you read it in as a df and need to convert to data.table is it still faster?

```{r echo=TRUE, include=TRUE}

tictoc::tic("data.table way")
dt <- as.data.table(df)
setkey(dt, state, date)
dt[, .(deaths_total = data.table::last(deaths)), by = state][order(-deaths_total)]
tictoc::toc()

```

## What if I like pipes, can data.table do that?

data.table can chain operations

```{r echo=TRUE, include=TRUE}


dt[, .(deaths_total = data.table::last(deaths)), by = state][order(-deaths_total)][deaths_total <= 100,]

# or you can put each on new line so long as the closing and opening brackets are together

dt[
  , .(deaths_total = data.table::last(deaths)), by = state
  ][
    order(-deaths_total)
  ][
    deaths_total <= 100,
  ]

# or you can use the pipe with .

dt[, .(deaths_total = data.table::last(deaths)), by = state] %>% 
  .[order(-deaths_total)] %>% 
  .[deaths_total <= 100]


```

## Adding new columns in place with := writes over the original object

so no assignment necessary (but also, be careful!)  

**data.table way**  

```{r echo=TRUE, include=TRUE}

# dplyr also has a last() function so I need to specify which package I want to use

dt[, state_death_total := data.table::last(deaths), by = state]

names(dt)

```

**tidyverse way**  

does not overwrite objects in place so you need to assign into a new object  

```{r echo=TRUE, include=TRUE}

# tidyverse way
df <- df %>% 
  group_by(state) %>% 
  mutate(state_death_total = last(deaths))

names(df)

```

## Walk through wrangling real data with data.table

### State-level data first

```{r echo=TRUE, include=TRUE}

#---- prep states ----

# Get state names for ones with more than 19 deaths
# if you want the total per group, but the group entries are
# cumulative, then you only want the last entry per group
# .N as an index on .SD, will give you the last row, then by = state
# then filter for deaths 20 or more
# and keep only the state names
state_totals <- states[, .SD[.N], by = state]

over19 <- state_totals[deaths >= 20, state]

# states a user can choose from
states <- states[state %chin% over19,]
state_choices <- states[, unique(state)]

# make new columns to show new cases and new deaths
count_cols = c('cases', 'deaths')
states[ , paste0('new_', count_cols) := lapply(.SD, function(x) x - data.table::shift(x, n = 1L, type = "lag")),
        by = state, .SDcols = count_cols]

# preview first 6 rows
states[, head(.SD)]

```

```{r echo=TRUE, include=TRUE}

date_range <- states[, range(unique(date))]
names(date_range) <- c("first", "last")

# rolling averages for new cases and new deaths
# using := as a function is the dt way to create multiple new columns at once,
# like using multiple mutate() in tidyverse
states <- states[
  , `:=`(
    nc_avg = frollmean(new_cases, 7L, align = "right"),
    nd_avg = frollmean(new_deaths, 7L, align = "right")
  )
  , by = state
][
  , c("nc_avg", "nd_avg") := lapply(.SD, ceiling), .SDcols = c("nc_avg", "nd_avg")
]

# preview first 6 rows
states[, head(.SD)]

```

### Then US-level data

```{r echo=TRUE, include=TRUE}

#---- prep US ----

# add column for new deaths
us[ , paste0('new_', count_cols) := lapply(.SD, function(x) x - data.table::shift(x, n = 1L, type = "lag")), .SDcols = count_cols]

# add rolling means
us[, `:=`(nd_avg = frollmean(new_deaths, 7L, align = "right"),
          deaths_avg = frollmean(deaths, 7L, align = "right"))]
# fill NA with 0
setnafill(us, type = "const", fill = 0L, cols=c("deaths_avg","nd_avg"))
# round up and convert to integer
us[, `:=`(nd_avg = as.integer(ceiling(nd_avg)),
          deaths_avg = as.integer(ceiling(deaths_avg)))]

us[, head(.SD)]

```

# Now use dtplyr if you want to do all of this with tidyverse syntax!

[dtplyr](https://dtplyr.tidyverse.org/)  
